---
output:
  md_document:
    variant: gfm
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
options(scipen=999)
options(digits=3)
set.seed(69839)
```

# segregation

[![CRAN Version](https://www.r-pkg.org/badges/version/segregation)](https://CRAN.R-project.org/package=segregation)
[![R build status](https://github.com/elbersb/segregation/workflows/R-CMD-check/badge.svg)](https://github.com/elbersb/segregation/actions)
[![Coverage status](https://codecov.io/gh/elbersb/segregation/branch/master/graph/badge.svg)](https://codecov.io/github/elbersb/segregation?branch=master)


An R package to calculate and decompose entropy-based, multigroup segregation indices, with a focus
on the Mutual Information Index (M) and Theil's Information Index (H).

Find more information in the [documentation](https://elbersb.de/segregation).

- calculate total, between, within, and local segregation
- decompose differences in total segregation over time
- estimate standard errors via bootstrapping
- every method returns a [tidy](https://vita.had.co.nz/papers/tidy-data.html) [data.table](https://rdatatable.gitlab.io/data.table/) for easy post-processing and plotting
- it's fast, because it uses the [`data.table`](https://rdatatable.gitlab.io/data.table/) package internally

Most of the procedures implemented in this package are described in more detail [in this working paper](https://osf.io/preprints/socarxiv/ya7zs/).

## Usage

The package provides an easy way to calculate segregation measures, based on the Mutual Information Index (M) and Theil's Entropy Index (H).

```{r}
library(segregation)

# example dataset with fake data provided by the package
mutual_total(schools00, "race", "school", weight = "n")
```

Standard errors in all functions can be estimated via boostrapping. This will also
apply bias-correction to the estimates:

```{r}
mutual_total(schools00, "race", "school", weight = "n", se = TRUE)
```

Decompose segregation into a between-state and a within-state term
(the sum of these equals total segregation):

```{r}
# between states
mutual_total(schools00, "race", "state", weight = "n")

# within states
mutual_total(schools00, "race", "school", within = "state", weight = "n")
```

Local segregation (`ls`) is a decomposition by units (here racial groups).
The sum of the proportion-weighted local segregation scores equals M:

```{r}
(local <- mutual_local(schools00, group = "school", unit = "race", weight = "n",
             se = TRUE, wide = TRUE))

sum(local$p * local$ls)
```

Decompose the difference in M between 2000 and 2005,
using iterative proportional fitting (IPF) and the Shapley decomposition,
as suggested by Karmel and Maclachlan (1988) and Deutsch et al. (2006):

```{r}
mutual_difference(schools00, schools05, group = "race", unit = "school",
                  weight = "n", method = "shapley")
```

Find more information in the [documentation](https://elbersb.de/segregation).

## How to install

To install the package from CRAN, use

```{r eval=FALSE}
install.packages("segregation")
```

To install the development version, use

```{r eval=FALSE}
devtools::install_github("elbersb/segregation")
```

## Papers using the Mutual information index

(list incomplete)

DiPrete, T. A., Eller, C. C., Bol, T., & van de Werfhorst, H. G. (2017). School-to-Work Linkages in the United States, Germany, and France. American Journal of Sociology, 122(6), 1869-1938. https://doi.org/10.1086/691327

Forster, A. G., & Bol, T. (2017). Vocational education and employment over the life course using a new measure of occupational specificity. Social Science Research, 70, 176-197. https://doi.org/10.1016/j.ssresearch.2017.11.004

Van Puyenbroeck, T., De Bruyne, K., & Sels, L. (2012). More than ‘Mutual Information’: Educational and sectoral gender segregation and their interaction on the Flemish labor market. Labour Economics, 19(1), 1-8. https://doi.org/10.1016/j.labeco.2011.05.002

Mora, R., & Ruiz-Castillo, J. (2003). Additively decomposable segregation indexes. The case of gender segregation by occupations and human capital levels in Spain. The Journal of Economic Inequality, 1(2), 147-179. https://doi.org/10.1023/A:1026198429377

## References on entropy-based segregation indices

Deutsch, J., Flückiger, Y. & Silber, J. (2009).
      Analyzing Changes in Occupational Segregation: The Case of Switzerland (1970–2000), in: Yves Flückiger, Sean F. Reardon, Jacques Silber (eds.) Occupational and Residential Segregation (Research on Economic Inequality, Volume 17), 171–202.

Elbers, B. (2019). A Method for Studying Difference in Segregation Levels Across Time and Space. SocArXiv Working Paper. https://osf.io/preprints/socarxiv/ya7zs/

Theil, H. (1971). Principles of Econometrics. New York: Wiley.

Frankel, D. M., & Volij, O. (2011). Measuring school segregation. Journal of Economic Theory, 146(1), 1-38. https://doi.org/10.1016/j.jet.2010.10.008

Mora, R., & Ruiz-Castillo, J. (2009). The Invariance Properties of the Mutual Information Index of Multigroup Segregation, in: Yves Flückiger, Sean F. Reardon, Jacques Silber (eds.) Occupational and Residential Segregation (Research on Economic Inequality, Volume 17), 33-53.

Mora, R., & Ruiz-Castillo, J. (2011). Entropy-based Segregation Indices. Sociological Methodology, 41(1), 159–194. https://doi.org/10.1111/j.1467-9531.2011.01237.x

Karmel, T. & Maclachlan, M. (1988). Occupational Sex Segregation — Increasing or Decreasing? Economic Record 64: 187-195. https://doi.org/10.1111/j.1475-4932.1988.tb02057.x

Watts, M. The Use and Abuse of Entropy Based Segregation Indices. Working Paper. URL: http://www.ecineq.org/ecineq_lux15/FILESx2015/CR2/p217.pdf
