---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
options(scipen=999)
options(digits=3)
set.seed(69839)
```

# segregation

An R package to calculate entropy-based segregation indices, with a focus
on the Mutual Information Index (M).

- calculate total, between, within, and local segregation
- decompose differences in total segregation over time
- estimate standard errors via bootstrapping
- every method returns a [tidy](http://vita.had.co.nz/papers/tidy-data.html) data frame (or [tibble](http://tibble.tidyverse.org), if the package is installed) for easy post-processing and plotting
- it's fast, because it uses the [`data.table`](https://github.com/Rdatatable/data.table/wiki) package internally

## Usage

The package provides an easy way to calculate total and local segregation, based on the Mutual Information Index.

```{r}
library(segregation)
# example dataset with fake data provided by the package
mutual_total(schools00, "race", "school", weight = "n")
```

Standard errors can be estimated via boostrapping:

```{r}
mutual_total(schools00, "race", "school", weight = "n", se = TRUE)
```

Local segregation (`ls`) by units (here racial groups), with standard errors:

```{r}
mutual_local(schools00, group="school", unit="race", weight = "n", se = TRUE)
```

Decompose the difference in segregation between 2000 and 2005, 
using the method developed by Mora and Ruiz-Castillo (2009):

```{r}
mutual_difference(schools00, schools05,
                  group="race", unit="school",
                  weight = "n", method = "mrc")
```

## How to install

To install the package from CRAN, use

```{r eval=FALSE}
install.packages("segregation") 
```

To install the development version, use

```{r eval=FALSE}
devtools::install_github("elbersb/segregation") 
```

To access the documentation ([PDF](https://cran.r-project.org/web/packages/segregation/segregation.pdf)), type

```{r eval=FALSE}
?segregation
```

## Papers using the Mutual information index

(list incomplete)

- DiPrete, T. A., Eller, C. C., Bol, T., & van de Werfhorst, H. G. (2017). School-to-Work Linkages in the United States, Germany, and France. American Journal of Sociology, 122(6), 1869-1938. https://doi.org/10.1086/691327
- Forster, A. G., & Bol, T. (2017). Vocational education and employment over the life course using a new measure of occupational specificity. Social Science Research, 70, 176-197. https://doi.org/10.1016/j.ssresearch.2017.11.004
- Van Puyenbroeck, T., De Bruyne, K., & Sels, L. (2012). More than ‘Mutual Information’: Educational and sectoral gender segregation and their interaction on the Flemish labor market. Labour Economics, 19(1), 1-8. https://doi.org/10.1016/j.labeco.2011.05.002
- Mora, R., & Ruiz-Castillo, J. (2003). Additively decomposable segregation indexes. The case of gender segregation by occupations and human capital levels in Spain. The Journal of Economic Inequality, 1(2), 147-179. https://doi.org/10.1023/A:1026198429377

## References on entropy-based segregation indices

Theil, Henri. (1971). Principles of Econometrics. New York: Wiley.

Frankel, D. M., & Volij, O. (2011). Measuring school segregation. Journal of Economic Theory, 146(1), 1-38. https://doi.org/10.1016/j.jet.2010.10.008

Mora, R., & Ruiz-Castillo, J. (2009). The Invariance Properties of the
   Mutual Information Index of Multigroup Segregation. Research on Economic Inequality, 17, 33-53.

Mora, R., & Ruiz-Castillo, J. (2011). Entropy-based Segregation Indices. Sociological Methodology, 41(1), 159–194. https://doi.org/10.1111/j.1467-9531.2011.01237.x
